---
title: "Introduction to summarytabl"
output: rmarkdown::html_vignette
description: >
  This document introduces you to some of summarytabl's most frequently used functions, and shows you how to apply them to data frames.
vignette: >
  %\VignetteIndexEntry{Introduction to summarytabl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the `summarytabl` package. This package creates simple summary tables for continuous, ordinal, and categorical data. If you notice a bug or have trouble using a function, please contact [Ama Nyame-Mensah](mailto:ama@anyamemensah.com), the package's maintainer.

To begin, load `summarytabl`

```{r setup}
library(summarytabl)
```

## Categorical variables

### Summarize a single categorical variable

You can generate a frequency table for a categorical variable using `cat_tbl()`

```{r}
cat_tbl(data = nlsy, var = "race")
```

You can also ignore specific values and remove missing values

```{r}
cat_tbl(data = nlsy, 
        var = "race",
        ignore = "Black",
        na.rm = TRUE)
```

### Summarize a categorical variable grouped by grouped another variable

A grouped frequency table for two categorical variables can be generated using `cat_group_tbl()`

```{r}
cat_group_tbl(data = nlsy,
              row_var = "gender",
              col_var = "bthwht")
```        

Like `cat_tbl()`, you can also specify whether to ignore specific values and remove missing values:

```{r}
cat_group_tbl(data = nlsy,
              row_var = "race",
              col_var = "bthwht",
              na.rm.row_var = TRUE,
              na.rm.col_var = FALSE,
              ignore = c(race = "Non-Black,Non-Hispanic"),
              pivot = "wider",
              only = NULL)
```    


## Multiple response variables

### Generate a table for a series of multiple response variables

You can generate a summary table for binary (e.g., Unselected/Selected), multiple-response (e.g., never, sometimes, often), and ordinal-scale (e.g., strongly disagree to strongly agree) variables using `select_tbl()`:

```{r}
select_tbl(data = depressive, var_stem = "dep")
```

You can also ignore specific values and remove missing values using the `ignore` argument:

```{r}
select_tbl(data = depressive, 
           var_stem = "dep", 
           ignore = 3)
```

To pivot the resulting table into the wider format, set the `pivot` argument to `wider`.

```{r}
select_tbl(data = depressive, 
           var_stem = "dep",
           pivot = "wider")
```

### Summarize a series of multiple response variables

A grouped frequency table for a series of multiple response variables with the same variable stem can be generated using `select_group_tbl()`

```{r}

dep_recoded <- 
  depressive |>
  dplyr::mutate(
    race = dplyr::case_match(.x = race,
                             1 ~ "Hispanic", 
                             2 ~ "Black", 
                             3 ~ "Non-Black/Non-Hispanic",
                             .default = NA)
  ) |>
  dplyr::mutate(
    dplyr::across(
      .cols = dplyr::starts_with("dep"),
      .fns = ~ dplyr::case_when(.x == 1 ~ "often", 
                           .x == 2 ~ "sometimes", 
                           .x == 3 ~ "hardly ever")
    )
  )

select_group_tbl(data = dep_recoded, 
                 var_stem = "dep",
                 group = "race",
                 pivot = "longer")
```        

Similar to `cat_group_tbl()`, you can indicate whether to exclude certain values and omit missing values. In this case, however, you should use the `var_stem` that you supply to the function when determining which values to exclude and whether to eliminate missing values:

```{r}
select_group_tbl(data = dep_recoded, 
                 var_stem = "dep",
                 group = "race",
                 pivot = "longer",
                 na_removal = "listwise",
                 ignore = c(dep = "often", race = "Non-Black/Non-Hispanic"))
```    

Use a list when you want to ignore multiple values from the same `var_stem` or `group` variable:

```{r}

select_group_tbl(data = dep_recoded, 
                 var_stem = "dep",
                 group = "race",
                 pivot = "longer",
                 na_removal = "listwise",
                 ignore = list(race = c("Hispanic", "Non-Black/Non-Hispanic")))
``` 

## Continuous variables

### Generate a descriptive statistics summary table for a series of continuous variables

A table presenting descriptive statistics (mean, sd, minimum, maximum, number of non-missing observations) for a series of continuous variables with the same variable stem can be generated using `mean_tbl()`

```{r}
mean_tbl(data = social_psy_data, 
         var_stem = "belong",
         na_removal = "pairwise")
``` 

Like the rest of the variables in the package, you can specify whether to ignore specific values from all variables associated with the variable stem.

```{r}
mean_tbl(data = social_psy_data, 
         var_stem = "belong",
         na_removal = "pairwise", 
         ignore = 3)
``` 

Applying variable labels can help make variable names more understandable.

```{r}
mean_tbl(data = social_psy_data, 
         var_stem = "belong",
         na_removal = "pairwise",
         var_labels = c(
          belong_1 = "I feel like I belong at this institution",
          belong_2 = "I feel like part of the community",
          belong_3 = "feel valued by this institution")
         )
``` 

### Generate a grouped table for a series of continuous variables

A grouped summary statistics table for a series of conitnuous variables with the same variable stem can be generated using `mean_group_tbl()`.

```{r}
mean_group_tbl(data = stem_social_psych, 
               var_stem = "belong_belong",
               group = "urm",
               group_type = "variable",
               na_removal = "pairwise")
``` 

Similar to `mean_tbl()`, you can indicate whether to exclude certain values from either the variables associated with `var_stem` or the group if `group_typ` is set to `variable`.

```{r}
mean_group_tbl(data = stem_social_psych, 
               var_stem = "belong_belong",
               group = "urm",
               group_type = "variable",
               na_removal = "pairwise",
               ignore = c(belong_belong = 3, urm = 0)
               )
``` 

To ignore multiple values, use a list:

```{r}
mean_group_tbl(data = stem_social_psych, 
               var_stem = "belong_belong",
               group = "urm",
               group_type = "variable",
               na_removal = "pairwise",
               ignore = list(belong_belong = c(1,2), urm = 0)
               )
``` 

With `mean_group_tbl()`, you can also extract patterns from variables in your dataset to serve as a grouping variable. A common application is examining summary statistics for a variable measured across several time points.

```{r}
mean_group_tbl(data = stem_social_psych, 
               var_stem = "belong_belong",
               group = "_w\\d",
               group_type = "pattern",
               na_removal = "pairwise")
```

If your grouping variable's name isn't sufficiently descriptive, or if `group` isn't clear enough, use the `group_name` argument to provide a more meaningful label for the resulting table.

```{r}
mean_group_tbl(data = stem_social_psych, 
               var_stem = "belong_belong",
               group = "_w\\d",
               group_type = "pattern",
               na_removal = "pairwise",
               group_name = "wave")
``` 

Finally, applying variable labels can help make variable names more understandable.

```{r}
mean_group_tbl(data = stem_social_psych, 
               var_stem = "belong_belong",
               group = "_w\\d",
               group_type = "pattern",
               na_removal = "pairwise",
               group_name = "wave",
               var_labels = c(
                 belong_belongStem_w1 = "I feel like I belong in computing",
                 belong_belongStem_w2 = "I feel like I belong in computing")
               )
``` 


